Loop-Aware Optimizations in PyPy's Tracing JIT

### 2012-08-17 talk/dls2012/paper.tex
tune plot
Update.

\begin{figure}
\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$L_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = $i_{2} + i_{3}$
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)
jump($L_0$, $p_{0}$, $p_{5}$)
\end{lstlisting}
\caption{An Unoptimized Trace of the example interpreter}
\label{fig:unopt-trace}
\end{figure}

@@

$L_1$($p_{0}$, $i_{3}$, $i_{4}$):
    $i_{8}$ = $i_{4}+i_{3}$
jump($L_1$, $p_{0}$, $i_{3}$, $i_8$)
\end{lstlisting}
\caption{The fully optimized loop of the example interpreter}
\label{fig:opt-trace}
\end{figure}

@@

\begin{figure*}
\begin{center}
{\smaller
\begin{tabular}{|l|r|r|r|r|r|r|r|}
\hline
 & CPython & PyPy  & PyPy & LuaJIT & LuaJIT & GCC \\
 &         & no LP &      & no LP  &        & -O3 \\
\hline
FFT(1024,32768) & 469.07 & 20.83 $\pm$ 0.039 & 12.73 $\pm$ 0.029 & 4.42 $\pm$ 0.017 & 2.76 $\pm$ 0.017 & 1.40 $\pm$ 0.082\\
\hline
FFT(1048576,2) & 58.93 & 4.12 $\pm$ 0.020 & 2.05 $\pm$ 0.007 & 1.33 $\pm$ 0.028 & 1.08 $\pm$ 0.014 & 0.83 $\pm$ 0.044\\
\hline
LU(100,4096) & 1974.14 & 32.22 $\pm$ 0.281 & 13.39 $\pm$ 0.063 & 8.57 $\pm$ 0.012 & 1.52 $\pm$ 0.014 & 1.33 $\pm$ 0.070\\
\hline
LU(1000,2) & 955.31 & 14.98 $\pm$ 0.436 & 5.99 $\pm$ 0.416 & 3.99 $\pm$ 0.014 & 0.67 $\pm$ 0.010 & 0.65 $\pm$ 0.077\\
\hline
MonteCarlo(268435456) & 618.89 & 20.60 $\pm$ 0.097 & 15.33 $\pm$ 0.163 & 3.92 $\pm$ 0.016 & 2.82 $\pm$ 0.009 & 1.69 $\pm$ 0.096\\
\hline
SOR(100,32768) & 1458.12 & 8.24 $\pm$ 0.002 & 2.66 $\pm$ 0.002 & 2.02 $\pm$ 0.009 & 1.31 $\pm$ 0.010 & 1.76 $\pm$ 0.088\\
\hline
SOR(1000,256) & 1210.45 & 6.48 $\pm$ 0.007 & 2.10 $\pm$ 0.005 & 1.63 $\pm$ 0.009 & 1.08 $\pm$ 0.010 & 1.49 $\pm$ 0.042\\
\hline
SparseMatMult(1000,5000,262144) & 371.66 & 24.25 $\pm$ 0.074 & 16.52 $\pm$ 0.077 & 9.64 $\pm$ 0.032 & 4.53 $\pm$ 0.032 & 1.84 $\pm$ 0.061\\
\hline
SparseMatMult(100000,1000000,1024) & 236.93 & 17.01 $\pm$ 0.025 & 8.75 $\pm$ 0.149 & 7.19 $\pm$ 0.016 & 2.42 $\pm$ 0.010 & 1.20 $\pm$ 0.053\\
\hline
\hline
conv3(1e5) & 50.14 & 1.09 $\pm$ 0.022 & 0.49 $\pm$ 0.028 & 0.67 $\pm$ 0.010 & 0.12 $\pm$ 0.010 & 0.52 $\pm$ 0.084\\
\hline
conv3(1e6) & 49.20 & 1.13 $\pm$ 0.043 & 0.51 $\pm$ 0.008 & 0.70 $\pm$ 0.008 & 0.18 $\pm$ 0.000 & 0.60 $\pm$ 0.064\\
\hline
conv3x3(1000,1000) & 138.95 & 0.70 $\pm$ 0.007 & 0.20 $\pm$ 0.009 & 0.49 $\pm$ 0.010 & 0.09 $\pm$ 0.010 & 0.17 $\pm$ 0.079\\
\hline
conv3x3(1000000,3) & 139.81 & 0.70 $\pm$ 0.005 & 0.21 $\pm$ 0.006 & 0.53 $\pm$ 0.008 & 0.13 $\pm$ 0.009 & 0.19 $\pm$ 0.061\\
\hline
conv5(1e5) & 74.65 & 1.22 $\pm$ 0.009 & 0.64 $\pm$ 0.005 & 0.84 $\pm$ 0.006 & 0.17 $\pm$ 0.010 & 0.55 $\pm$ 0.047\\
\hline
conv5(1e6) & 77.94 & 1.26 $\pm$ 0.009 & 0.68 $\pm$ 0.014 & 0.87 $\pm$ 0.010 & 0.21 $\pm$ 0.013 & 0.58 $\pm$ 0.049\\
\hline
dilate3x3(1000,1000) & 137.52 & 4.35 $\pm$ 0.014 & 3.91 $\pm$ 0.037 & 0.48 $\pm$ 0.014 & 0.09 $\pm$ 0.006 & 0.17 $\pm$ 0.061\\
\hline
sobel(1000,1000) & 104.02 & 0.49 $\pm$ 0.009 & 0.21 $\pm$ 0.004 & 0.60 $\pm$ 0.012 & 0.24 $\pm$ 0.018 & 0.17 $\pm$ 0.061\\
\hline
sqrt(float) & 14.99 & 1.37 $\pm$ 0.001 & 0.89 $\pm$ 0.000 & 1.06 $\pm$ 0.013 & 0.83 $\pm$ 0.010 & 0.85 $\pm$ 0.088\\
\hline
sqrt(int) & 13.91 & 3.22 $\pm$ 0.033 & 2.65 $\pm$ 0.001 & - & - & 1.25 $\pm$ 0.053\\
\hline
sqrt(Fix16) & 463.46 & 5.12 $\pm$ 0.005 & 2.96 $\pm$ 0.007 & - & - & 1.34 $\pm$ 0.061\\
\hline
\end{tabular}
}
\end{center}
\label{fig:benchmarks}
\caption{Benchmark results in seconds with 95\% confidence intervals. The leftmost column gives the
name of each benchmark and the values of the benchmark parameters used. The different benchmarks and the meaning of their parameters are described in Section~\ref{sec:benchmarks}.}
\end{figure*}

@@

The benchmarks are
\begin{itemize}
\item {\bf sqrt}$\left(T\right)$: approximates the square root of $y$. The approximation is 
initialized to $x_0=y/2$ and the benchmark consists of a single loop updating this
approximation using $x_i = \left( x_{i-1} + y/x_{i-1} \right) / 2$ for $1\leq i < 10^8$. 
Only the latest calculated value $x_i$ is kept alive as a local variable within the loop.
There are three different versions of this benchmark where $x_i$
  is represented with different type of objects, $T$,: int's, float's and
  Fix16's. The latter, Fix16, is a custom class that implements
  fixpoint arithmetic with 16 bits precision. In Python there is only
  a single implementation of the benchmark that gets specialized
  depending on the class of it's input argument, $y$, while in C,
  there are three different implementations. In Lua there is no support for
  integers so only the floating point number is provided.
\item {\bf conv3}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $3$. A single loop
is used to calculate a vector ${\bf b} = \left(b_1, \cdots, b_{n-2}\right)$ from a vector
${\bf a} = \left(a_1, \cdots, a_n\right)$ and a kernel ${\bf k} = \left(k_1, k_2, k_3\right)$ using 
$b_i = k_3 a_i + k_2 a_{i+1} + k_1 a_{i+2}$ for $1 \leq i \leq n-2$. Both the output vector, $\bf b$, 
and the input vectors, $\bf a$ and $\bf k$, are allocated prior to running the benchmark. It is executed 
with $n=10^5$ and $n=10^6$.
\item {\bf conv5}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $5$. Similar to conv3, but with 
${\bf k} = \left(k_1, k_2, k_3, k_4, k_5\right)$. The enumeration of the elements in $\bf k$ is still 
hardcoded into the implementation making the benchmark consist of a single loop too.
\item {\bf conv3x3}$\left(n,m\right)$: two-dimensional convolution with kernel of fixed
  size $3 \times 3$ using a custom class to represent two-dimensional
  arrays. It is implemented as two nested loops that iterates over the elements of the 
$m\times n$ output matrix ${\bf B} = \left(b_{i,j}\right)$ and calculates each element from the input matrix
${\bf A} = \left(a_{i,j}\right)$ and a kernel ${\bf K} = \left(k_{i,j}\right)$ using $b_{i,j} = $
\begin{equation}
  \label{eq:convsum}
  \begin{array}{lclclc}
    k_{3,3} a_{i-1,j-1} &+& k_{3,2} a_{i-1,j} &+& k_{3,1} a_{i-1,j+1} & + \\
    k_{2,3} a_{i,j-1}   &+& k_{2,2} a_{i,j}   &+& k_{2,1} a_{i,j+1}   & + \\
    k_{1,3} a_{i+1,j-1} &+& k_{1,2} a_{i+1,j} &+& k_{1,1} a_{i+1,j+1}  \\
  \end{array}
\end{equation}
for $2 \leq i \leq m-1$ and $2 \leq j \leq n-1$.
The memory for storing the matrices are again allocated outside the benchmark and $(n,m)=(1000,1000)$ 
as well as $(n,m)=(1000000,3)$ was used.
\item {\bf dilate3x3}$\left(n\right)$: two-dimensional dilation with kernel of fixed
  size $3 \times 3$. This is similar to convolution but instead of
  summing over the terms in Equation~\ref{eq:convsum}, the maximum over those terms is taken. That places a
  external call to a max function within the loop that prevents some
  of the optimizations.
\item {\bf sobel}$\left(n\right)$: a low-level video processing algorithm used to
  locate edges in an image. It calculates the gradient magnitude
  using sobel derivatives. A Sobel x-derivative, $D_x$, of a $n \times n$ image, ${I}$, is formed
by convolving ${I}$ with
\begin{equation}
  {K} = \left(
  \begin{array}{ccc}
    -1 & 0 & 1 \\
    -2 & 0 & 2 \\
    -1 & 0 & 1 \\
  \end{array}
  \right) ,
\end{equation}
and a Sobel y-derivative, $D_y$, is formed convolving $I$ with $K^\top$. The gradient magnitude is 
then formed for each pixel independently by $\sqrt{D_x^2 + D_y^2}$. The two convolutions and the pixelwise
magnitude calculation are combined in the implementation of this benchmark and calculated in a single pass over
the input image. This single pass consists of two nested loops with a somewhat larger amount of calculations 
performed each iteration as compared to the other benchmarks.
\end{itemize}

>>>>

\begin{figure}
\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$L_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = $i_{2} + i_{3}$
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)
jump($L_0$, $p_{0}$, $p_{5}$)
\end{lstlisting}
\caption{An Unoptimized Trace of the Example Interpreter}
\label{fig:unopt-trace}
\end{figure}

@@

$L_1$($p_{0}$, $i_{3}$, $i_{4}$):
    $i_{8}$ = $i_{4}+i_{3}$
jump($L_1$, $p_{0}$, $i_{3}$, $i_8$)
\end{lstlisting}
\caption{The fully optimized loop of the Example Interpreter}
\label{fig:opt-trace}
\end{figure}

@@

\begin{figure*}
\begin{center}
{\smaller
\begin{tabular}{|l|r|r|r|r|r|r|r|}
\hline
 & CPython & PyPy  & PyPy & LuaJIT & LuaJIT & GCC \\
 &         & no LP &      & no LP  &        & -O3 \\
\hline
FFT(1024,32768) & 469.07 & 20.83 $\pm$ 0.039 & 12.73 $\pm$ 0.029 & 4.42 $\pm$ 0.017 & 2.76 $\pm$ 0.017 & 1.40 $\pm$ 0.082\\
\hline
FFT(1048576,2) & 58.93 & 4.12 $\pm$ 0.020 & 2.05 $\pm$ 0.007 & 1.33 $\pm$ 0.028 & 1.08 $\pm$ 0.014 & 0.83 $\pm$ 0.044\\
\hline
LU(100,4096) & 1974.14 & 32.22 $\pm$ 0.281 & 13.39 $\pm$ 0.063 & 8.57 $\pm$ 0.012 & 1.52 $\pm$ 0.014 & 1.33 $\pm$ 0.070\\
\hline
LU(1000,2) & 955.31 & 14.98 $\pm$ 0.436 & 5.99 $\pm$ 0.416 & 3.99 $\pm$ 0.014 & 0.67 $\pm$ 0.010 & 0.65 $\pm$ 0.077\\
\hline
MonteCarlo(268435456) & 618.89 & 20.60 $\pm$ 0.097 & 15.33 $\pm$ 0.163 & 3.92 $\pm$ 0.016 & 2.82 $\pm$ 0.009 & 1.69 $\pm$ 0.096\\
\hline
SOR(100,32768) & 1458.12 & 8.24 $\pm$ 0.002 & 2.66 $\pm$ 0.002 & 2.02 $\pm$ 0.009 & 1.31 $\pm$ 0.010 & 1.76 $\pm$ 0.088\\
\hline
SOR(1000,256) & 1210.45 & 6.48 $\pm$ 0.007 & 2.10 $\pm$ 0.005 & 1.63 $\pm$ 0.009 & 1.08 $\pm$ 0.010 & 1.49 $\pm$ 0.042\\
\hline
SparseMatMult(1000,5000,262144) & 371.66 & 24.25 $\pm$ 0.074 & 16.52 $\pm$ 0.077 & 9.64 $\pm$ 0.032 & 4.53 $\pm$ 0.032 & 1.84 $\pm$ 0.061\\
\hline
SparseMatMult(100000,1000000,1024) & 236.93 & 17.01 $\pm$ 0.025 & 8.75 $\pm$ 0.149 & 7.19 $\pm$ 0.016 & 2.42 $\pm$ 0.010 & 1.20 $\pm$ 0.053\\
\hline
%\hline
%conv3(1e5) & 50.14 & 1.09 $\pm$ 0.022 & 0.49 $\pm$ 0.028 & 0.67 $\pm$ 0.010 & 0.12 $\pm$ 0.010 & 0.52 $\pm$ 0.084\\
\hline
conv3(1e6) & 49.20 & 1.13 $\pm$ 0.043 & 0.51 $\pm$ 0.008 & 0.70 $\pm$ 0.008 & 0.18 $\pm$ 0.000 & 0.60 $\pm$ 0.064\\
\hline
conv3x3(1000,1000) & 138.95 & 0.70 $\pm$ 0.007 & 0.20 $\pm$ 0.009 & 0.49 $\pm$ 0.010 & 0.09 $\pm$ 0.010 & 0.17 $\pm$ 0.079\\
\hline
%conv3x3(1000000,3) & 139.81 & 0.70 $\pm$ 0.005 & 0.21 $\pm$ 0.006 & 0.53 $\pm$ 0.008 & 0.13 $\pm$ 0.009 & 0.19 $\pm$ 0.061\\
%\hline
%conv5(1e5) & 74.65 & 1.22 $\pm$ 0.009 & 0.64 $\pm$ 0.005 & 0.84 $\pm$ 0.006 & 0.17 $\pm$ 0.010 & 0.55 $\pm$ 0.047\\
%\hline
%conv5(1e6) & 77.94 & 1.26 $\pm$ 0.009 & 0.68 $\pm$ 0.014 & 0.87 $\pm$ 0.010 & 0.21 $\pm$ 0.013 & 0.58 $\pm$ 0.049\\
%\hline
dilate3x3(1000,1000) & 137.52 & 4.35 $\pm$ 0.014 & 3.91 $\pm$ 0.037 & 0.48 $\pm$ 0.014 & 0.09 $\pm$ 0.006 & 0.17 $\pm$ 0.061\\
\hline
sobel(1000,1000) & 104.02 & 0.49 $\pm$ 0.009 & 0.21 $\pm$ 0.004 & 0.60 $\pm$ 0.012 & 0.24 $\pm$ 0.018 & 0.17 $\pm$ 0.061\\
\hline
sqrt(float) & 14.99 & 1.37 $\pm$ 0.001 & 0.89 $\pm$ 0.000 & 1.06 $\pm$ 0.013 & 0.83 $\pm$ 0.010 & 0.85 $\pm$ 0.088\\
\hline
sqrt(int) & 13.91 & 3.22 $\pm$ 0.033 & 2.65 $\pm$ 0.001 & - & - & 1.25 $\pm$ 0.053\\
\hline
sqrt(Fix16) & 463.46 & 5.12 $\pm$ 0.005 & 2.96 $\pm$ 0.007 & - & - & 1.34 $\pm$ 0.061\\
\hline
\end{tabular}
}
\end{center}
\label{fig:benchmarks}
\caption{Benchmark results in seconds with 95\% confidence intervals. The leftmost column gives the
name of each benchmark and the values of the benchmark parameters used. The different benchmarks and the meaning of their parameters are described in Section~\ref{sec:benchmarks}.}
\end{figure*}

@@

The benchmarks are
\begin{itemize}
\item {\bf sqrt}$\left(T\right)$: approximates the square root of $y$. The approximation is 
initialized to $x_0=y/2$ and the benchmark consists of a single loop updating this
approximation using $x_i = \left( x_{i-1} + y/x_{i-1} \right) / 2$ for $1\leq i < 10^8$. 
Only the latest calculated value $x_i$ is kept alive as a local variable within the loop.
There are three different versions of this benchmark where $x_i$
  is represented with different type of objects, $T$,: int's, float's and
  Fix16's. The latter, Fix16, is a custom class that implements
  fixpoint arithmetic with 16 bits precision. In Python there is only
  a single implementation of the benchmark that gets specialized
  depending on the class of it's input argument, $y$, while in C,
  there are three different implementations. In Lua there is no support for
  integers so only the floating point number is provided.
\item {\bf conv3}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $3$. A single loop
is used to calculate a vector ${\bf b} = \left(b_1, \cdots, b_{n-2}\right)$ from a vector
${\bf a} = \left(a_1, \cdots, a_n\right)$ and a kernel ${\bf k} = \left(k_1, k_2, k_3\right)$ using 
$b_i = k_3 a_i + k_2 a_{i+1} + k_1 a_{i+2}$ for $1 \leq i \leq n-2$. Both the output vector, $\bf b$, 
and the input vectors, $\bf a$ and $\bf k$, are allocated prior to running the benchmark. It is executed 
with $n=10^5$.
%\item {\bf conv5}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $5$. Similar to conv3, but with 
%${\bf k} = \left(k_1, k_2, k_3, k_4, k_5\right)$. The enumeration of the elements in $\bf k$ is still 
hardcoded into the implementation making the benchmark consist of a single loop too.
\item {\bf conv3x3}$\left(n,m\right)$: two-dimensional convolution with kernel of fixed
  size $3 \times 3$ using a custom class to represent two-dimensional
  arrays. It is implemented as two nested loops that iterates over the elements of the 
$m\times n$ output matrix ${\bf B} = \left(b_{i,j}\right)$ and calculates each element from the input matrix
${\bf A} = \left(a_{i,j}\right)$ and a kernel ${\bf K} = \left(k_{i,j}\right)$ using $b_{i,j} = $
\begin{equation}
  \label{eq:convsum}
  \begin{array}{lclclc}
    k_{3,3} a_{i-1,j-1} &+& k_{3,2} a_{i-1,j} &+& k_{3,1} a_{i-1,j+1} & + \\
    k_{2,3} a_{i,j-1}   &+& k_{2,2} a_{i,j}   &+& k_{2,1} a_{i,j+1}   & + \\
    k_{1,3} a_{i+1,j-1} &+& k_{1,2} a_{i+1,j} &+& k_{1,1} a_{i+1,j+1}  \\
  \end{array}
\end{equation}
for $2 \leq i \leq m-1$ and $2 \leq j \leq n-1$.
The memory for storing the matrices are again allocated outside the benchmark and $(n,m)=(1000,1000)$ 
 was used.
\item {\bf dilate3x3}$\left(n\right)$: two-dimensional dilation with kernel of fixed
  size $3 \times 3$. This is similar to convolution but instead of
  summing over the terms in Equation~\ref{eq:convsum}, the maximum over those terms is taken. That places a
  external call to a max function within the loop that prevents some
  of the optimizations.
\item {\bf sobel}$\left(n\right)$: a low-level video processing algorithm used to
  locate edges in an image. It calculates the gradient magnitude
  using sobel derivatives. A Sobel x-derivative, $D_x$, of a $n \times n$ image, ${I}$, is formed
by convolving ${I}$ with
\begin{equation}
  {K} = \left(
  \begin{array}{ccc}
    -1 & 0 & 1 \\
    -2 & 0 & 2 \\
    -1 & 0 & 1 \\
  \end{array}
  \right) ,
\end{equation}
and a Sobel y-derivative, $D_y$, is formed convolving $I$ with $K^\top$. The gradient magnitude is 
then formed for each pixel independently by $\sqrt{D_x^2 + D_y^2}$. The two convolutions and the pixelwise
magnitude calculation are combined in the implementation of this benchmark and calculated in a single pass over
the input image. This single pass consists of two nested loops with a somewhat larger amount of calculations 
performed each iteration as compared to the other benchmarks.
\end{itemize}
